---
title: "laggar"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{laggar}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
  ---
  
```{r, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
```
  
```{r setup}
  library(laggar)
  library(sf)
  library(ggplot2)
  library(mgcv)
```
  
  # Simulate data
  
```{r}
  set.seed(100230)
boundary <- st_polygon(list(matrix(c(0,1000,1000,0,0,
0,0,1000,1000,0),ncol=2)))

adults <- st_sample(boundary, size = 300)

seedtraps <- data.frame(
x = rep(seq(50,950,100), 10),
y = rep(seq(50,950,100), each = 10)
) |>
st_as_sf(coords = c("x","y"), remove = FALSE)

ggplot() + geom_sf(data = boundary) + geom_sf(data = adults) +
geom_sf(data = seedtraps, colour = "red") +
theme_void()
```

Now simulate number of seeds by using number of trees within distance (ignoring boundary effects of polygon for this):

```{r}
vshort_distance <- sum_within_dist(seedtraps, adults, boundary, 10)
short_distance <- sum_within_dist(seedtraps, adults, boundary, 25)
medium_distance <- sum_within_dist(seedtraps, adults, boundary, 75)
long_distance <- sum_within_dist(seedtraps, adults, boundary, 125)
vlong_distance <- sum_within_dist(seedtraps, adults, boundary, 200)

total_pred <- -0.5*vshort_distance[[1]] + 1*short_distance[[1]] +
  0.2*medium_distance[[1]] + 
  0.05*long_distance[[1]] + 0.01*vlong_distance[[1]]

nseeds <- rpois(100, total_pred)
```


# Now do data prep

Establish sequences of distance of interest

```{r}
dist_seq <- seq(10,200,10)
```

```{r}

x = seedtraps
y = adults
dist_seq = dist_seq
poly = boundary

dist <- laggar:::.index_check(minindex = NULL, maxindex = NULL, 
                              incindex = NULL, index_seq = NULL)


sum_within_dist(x = x, y = y, bound_poly = bound_poly, dist = 20)

## sum the number of points within dist of each seedtrap
sumdist <- lapply(dist_seq, sum_within_dist, x = x, y = y, poly = poly)

## no idea what this does
num_points <- sapply(sumdist, "[[", 1) # extracts the first item in each sublist, i.e. npoints
num_points <- t(apply(num_points, 1, function(x) diff(c(0, 
                                                        x))))


if (!is.null(poly)) {
  buffer_area <- sapply(sumdist, "[[", 2)
  buffer_area <- t(apply(buffer_area, 1, function(x) diff(c(0, 
                                                            x))))
  points_parea <- num_points/buffer_area
} else {
  buffer_area <- NA
  points_parea <- NA
}
return(list(num_points = num_points, buffer_area = buffer_area, 
            points_parea = points_parea))



## original code from vignette...
points_parea <- lg_points(seedtraps, adults, 
                          dist_seq = dist_seq, bound_poly = boundary)
points_perha <- 1e4*points_parea$points_parea

```


Assemble into lagR object:

```{r}
lgr <- lg_assembly(response = nseeds, 
                   covariate = points_perha, 
                   index_seq = dist_seq)
str(lgr)


    index <- laggar:::.index_check(minindex = NULL, maxindex = NULL, 
        incindex = NULL, index_seq = index_seq)
    if (!identical(nrow(covariate), length(response))) 
        stop("number of rows of response must be equal to length of covariate")
    if (anyNA(covariate)) {
        cov_na_index <- apply(covariate, 1, anyNA)
        if (all(cov_na_index)) {
            stop("covariate matrix has NAs present in every row")
        }
        else {
            message("Removing rows where covariate matrix has NAs")
            cov_notna_index <- !cov_na_index
        }
    }
    else {
        cov_notna_index <- rep(TRUE, length(response))
    }
    if (anyNA(response)) {
        resp_na_index <- is.na(response)
        if (all(resp_na_index)) {
            stop("Response is entirely NAs")
        }
        else {
            message("Removing rows where response is NA")
            resp_notna_index <- !resp_na_index
        }
    }
    else {
        resp_notna_index <- rep(TRUE, length(response))
    }
    
    
    ## this ensures that the rows in both response and covariates match!
    notna_index <- (cov_notna_index + resp_notna_index) == 2
    response <- response[notna_index]
    covariate <- covariate[notna_index, ]
    index_matrix <- matrix(index, nrow = nrow(covariate), ncol = ncol(covariate), 
        byrow = TRUE)
    data_list <- list(response = response, covariate = covariate, 
        index = index_matrix)
    return(data_list)


```

Graphical checks:

```{r}
lg_plot(lgr)
```



# Model

```{r}
modfit <- gam(response ~ te(index, by = covariate, bs = "tp", k = 10),
              data = lgr,
              family = "poisson", method = "REML")
```

```{r}
summary(modfit)
```

```{r}
plot(modfit)
```

