---
title: "laggar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{laggar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The laggar R package is designed to assist in data preparation for scalar-on-function regression within the mgcv package. Scalar-on-function regression can also be termed zone-of-influence modelling, or modelling over lags, and has a variety of applications. For more information on the statistical method, see the preprint of Miller et al [2025](https://doi.org/10.48550/arXiv.2508.07915).

```{r}
# install laggar if not already installed
# remotes::install_github("NERC-CEH/laggar")
```


```{r setup}
library(laggar)
library(sf)
library(ggplot2)
library(mgcv)
```

# Points data

## Simulate data

```{r}
set.seed(100230)
boundary <- st_polygon(list(matrix(c(0,1000,1000,0,0,
                                     0,0,1000,1000,0),ncol=2)))
adults <- st_sample(boundary, size = 300)
seedtraps <- data.frame(
  x = rep(seq(50,950,100), 10),
  y = rep(seq(50,950,100), each = 10)
) |>
  st_as_sf(coords = c("x","y"), remove = FALSE)
ggplot() + geom_sf(data = boundary) + geom_sf(data = adults) +
  geom_sf(data = seedtraps, colour = "red") +
  theme_void()
```

Now simulate number of seeds by using number of trees within distance (ignoring boundary effects of polygon for this):

```{r}
vshort_distance <- sum_within_dist(seedtraps, adults, boundary, 10)
short_distance <- sum_within_dist(seedtraps, adults, boundary, 25)
medium_distance <- sum_within_dist(seedtraps, adults, boundary, 75)
long_distance <- sum_within_dist(seedtraps, adults, boundary, 125)
vlong_distance <- sum_within_dist(seedtraps, adults, boundary, 200)

total_pred <- -0.5*vshort_distance[[1]] + 1*short_distance[[1]] +
  0.2*medium_distance[[1]] + 
  0.05*long_distance[[1]] + 0.01*vlong_distance[[1]]

nseeds <- rpois(100, total_pred)
```


## Now do data prep

Establish sequences of distance of interest

```{r}
dist_seq <- seq(10,200,10)
```

```{r}
points_parea <- lg_points(seedtraps, adults, 
                          dist_seq = dist_seq, poly = boundary)
points_perha <- 1e4*points_parea$value_parea
```


Assemble into lagR object:

```{r}
lgr <- lg_assembly(response = nseeds, 
                   covariate = points_perha, 
                   index_seq = dist_seq)
```

Graphical checks:

```{r, fig.dim = c(7,5)}
lg_plot(lgr)
```



## Model

```{r}
modfit <- gam(response ~ te(index, by = covariate, bs = "tp", k = 10),
              data = lgr,
              family = "poisson", method = "REML")
```

```{r}
summary(modfit)
```

```{r}
plot(modfit)
```


# Raster data 

## Simulate data

Load in a couple of extra packages for data simulation and plotting:

```{r setup-2}
library(prioritizr)
library(tidyterra)
library(terra)
```

We'll use the sampling locations created above, for simplicity. But we'll now add a raster layer over the area of interest: 

```{r}
# background rainfall data
r <- terra::rast(vect(boundary), nrow = 100, ncol = 100, vals = 1, crs = "EPSG:32630")
renv <- simulate_data(x = r, n = 1, scale = 0.8, intensity = 22, sd = 1)
st_crs(seedtraps) <- "EPSG:32630"

ggplot() +
  geom_spatraster(data = renv) +
  geom_point(data = seedtraps, aes(x, y), colour = "black") +
  scale_fill_viridis_c() +
  theme_void()
```


Simulate a monotonically decreasing effect of the raster layer with increasing distance from the measurement point:

```{r}
val_distance <- values_within_dist_rast(seedtraps, renv, func = "mean",
                                        dist_seq = c(50,100,150,200),
                                        plot_doughnuts = FALSE)

total_pred <- 1*val_distance$value[,1] +
  0.3*val_distance$value[,2] + 
  0.05*val_distance$value[,3] + 
  0.01*val_distance$value[,3]

growth_rate <- rnorm(100, total_pred,0.25)

```


## Data prep

Establish sequences of distance of interest.

```{r}
dist_seq <- seq(20,200,20)
```

Extract the summary values over the sequence of distances:

```{r}
rast_sum <- lg_rast(x = seedtraps,
                    y = renv,
                    func = "mean",
                    dist_seq = dist_seq, mindist = NULL, maxdist = NULL, incdist = NULL,
                    plot_doughnuts = TRUE, object_n = c(1, 10, 20), bg_layer = renv)

```


Assemble into lagR object:

```{r}
lgr <- lg_assembly(response = growth_rate, 
                   covariate = rast_sum$value_parea, 
                   index_seq = dist_seq)
```

Graphical checks:

```{r, fig.dim = c(7,5)}
lg_plot(lgr)
```



## Model

```{r}
modfit <- gam(response ~ te(index, by = covariate, bs = "tp", k = 10),
              data = lgr, method = "REML")
```

```{r}
summary(modfit)
```

```{r}
plot(modfit)
```

