---
title: "Laggar raster"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  ---
  
  ```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

```{r}
# install laggar if not already installed
# remotes::install_github("NERC-CEH/laggar")
```


```{r setup}
library(prioritizr)
library(tidyterra)
library(laggar)
library(sf)
```

# Simulate spatial data

```{r}
set.seed(100230)
boundary <- st_polygon(list(matrix(c(0,1000,1000,0,0,
                                     0,0,1000,1000,0),ncol=2)))

adults <- st_sample(boundary, size = 300)

seedtraps <- data.frame(
  x = rep(seq(50,950,100), 10),
  y = rep(seq(50,950,100), each = 10)
) |>
  st_as_sf(coords = c("x","y"), remove = FALSE,
           crs = "EPSG:32630")

plants <- data.frame(
  x = round((runif(100,50,950)),0),
  y = round(rep(runif(100,50,950)),0)
) |>
  st_as_sf(coords = c("x","y"), remove = FALSE,
           crs = "EPSG:32630")
```

Now we need to simulate some background raster data, rainfall, for the area of interest:

```{r}
# background rainfall data
r <- rast(vect(boundary), nrow = 100, ncol = 100, vals = 1, crs = "EPSG:32630")
renv <- simulate_data(x = r, n = 1, scale = 0.8, intensity = 22, sd = 1)

ggplot() +
  geom_spatraster(data = renv) +
  geom_point(data = plants, aes(x, y), colour = "black") +
  scale_fill_viridis_c() +
  theme_void()

```


# Now do data prep

Establish sequences of distance of interest.

```{r}
dist_seq <- seq(10,200,50)
```

```{r}
rast_sum <- lg_rast(x = plants,
                    y = renv,
                    func = "sum",
                    dist_seq = dist_seq, mindist = NULL, maxdist = NULL, incdist = NULL,
                    plot_doughnuts = TRUE, object_n = 1)
```


Assemble into lagR object:

```{r}
lgr <- lg_assembly(response = plants, 
                   covariate = rast_sum$value_parea, 
                   index_seq = dist_seq)
```

Graphical checks:

```{r, fig.dim = c(7,5)}
lg_plot(lgr)
```



# Model

```{r}
modfit <- gam(response ~ te(index, by = covariate, bs = "tp", k = 10),
              data = lgr,
              family = "poisson", method = "REML")
```

```{r}
summary(modfit)
```

```{r}
plot(modfit)
```
