---
title: "Laggar raster"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{laggar raster}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  ---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
# install laggar if not already installed
# remotes::install_github("NERC-CEH/laggar")
```


```{r setup}
library(prioritizr)
library(tidyterra)
library(laggar)
library(sf)
library(terra)
library(ggplot2)
library(ggnewscale)
library(mgcv)
```

# Simulate spatial data

```{r}
set.seed(100230)
boundary <- st_polygon(list(matrix(c(0,1000,1000,0,0,
                                     0,0,1000,1000,0),ncol=2))) %>% 
  st_sfc(crs = "EPSG:32630")

adults <- st_sample(boundary, size = 300) %>% 
  st_sfc(crs = "EPSG:32630")

seedtraps <- data.frame(
  x = rep(seq(50,950,100), 10),
  y = rep(seq(50,950,100), each = 10)
) %>% 
  st_as_sf(coords = c("x","y"), remove = FALSE,
           crs = "EPSG:32630")
```

# simulate response and locations

```{r}
# simulate response and locations
vshort_distance <- sum_within_dist(seedtraps, adults, boundary, 10)
short_distance <- sum_within_dist(seedtraps, adults, boundary, 25)
medium_distance <- sum_within_dist(seedtraps, adults, boundary, 75)
long_distance <- sum_within_dist(seedtraps, adults, boundary, 125)
vlong_distance <- sum_within_dist(seedtraps, adults, boundary, 200)

total_pred <- -0.5*short_distance[[1]] + 1*short_distance[[1]] +
  0.2*medium_distance[[1]] + 
  0.05*long_distance[[1]] + 0.01*vlong_distance[[1]]

nseeds <- rpois(100, total_pred)

plants <- data.frame(
  nseeds = nseeds,
  x = round(runif(100,50,950),0),
  y = round(runif(100,50,950),0)
) %>% 
  st_as_sf(coords = c("x","y"), remove = FALSE,
           crs = "EPSG:32630")
```

Now we need to simulate some background raster data, rainfall, for the area of interest:

```{r}
# background rainfall data
r <- rast(vect(boundary), nrow = 100, ncol = 100, vals = 1, crs = "EPSG:32630")
renv <- simulate_data(x = r, n = 1, scale = 0.8, intensity = 22, sd = 1)

ggplot() +
  geom_spatraster(data = renv) +
  geom_point(data = plants, aes(x, y), colour = "black") +
  scale_fill_viridis_c() +
  theme_void()

```


# Now do data prep

Establish sequences of distance of interest.

```{r}
dist_seq <- seq(20,200,20)
```

```{r}
rast_sum <- lg_rast(x = plants,
                    y = renv,
                    func = "sum",
                    dist_seq = dist_seq, mindist = NULL, maxdist = NULL, incdist = NULL,
                    plot_doughnuts = TRUE, object_n = c(1, 10, 20), bg_layer = renv)

```


Assemble into lagR object:

```{r}
lgr <- lg_assembly(response = plants$nseeds, 
                   covariate = rast_sum$value_parea, 
                   index_seq = dist_seq)
```

Graphical checks:

```{r, fig.dim = c(7,5)}
lg_plot(lgr)
```



# Model

```{r}
modfit <- gam(response ~ te(index, by = covariate, bs = "tp", k = 10),
              data = lgr,
              family = "poisson", method = "REML")
```

```{r}
summary(modfit)
```

```{r}
plot(modfit)
```
